Presolving for\\ Mixed-Integer Semidefinite Optimization

### 2021-03-23 tex/MISDP-Presolving/MISDPpresolving.tex
add right hand side
correct bound tightening in presolving document

One can also use the methods in Section~\ref{sec:CoefficientTightening} to
compute bounds on each $y_k$ variable. For this we only need that all $A^k$
are psd, but $y_k$ does not need to be binary. We need the following:

@@

\todo[inline]{FM: If we tighten the bounds of a variable~$y_k$, we need to
  scale the linear constraints accordingly.}

@@

From the lemma, we conclude that one can tighten the upper bounds of all
variables $y_{\ell}$ to $\min\, \{u_{\ell}, \hat{\mu}_{\ell}\}$. This is
implemented, but has not tightened any bound in presolving.

@@

\begin{proof}
  The constraint matrix of $S^k$ with respect to the equations and the
  inequalities $x - z \leq 0$ can be written as follows:
  \[
    M \define
    \begin{pmatrix}
      1 & \dots & 1 & 0 & \dots & 0 \\
      0 & \dots & 0 & 1 & \dots & 1 \\
      1 &       &   & -1 &      &   \\
        & \ddots&   &    & \ddots & \\
      0 &       & 1 & 0 &      & -1
    \end{pmatrix}.
  \]
  This matrix is totally unimodular. This can be seen by multiplying the
  first row with $-1$. Then we have exactly one $+1$ and $-1$ in each
  column. Thus, $M$ is the node-arc indicence matrix of a directed graph
  and hence TU. To take care of the variables bounds, one would add
  identity matrices to $M$, which do not change the property of being TU.
\end{proof}

@@

  Then Aghezzaf and Wolsey consider lot sizing problems for given demands $D_1,
  \dots, D_n$:
  \begin{align*}
    \min\quad & c\T x + d\T z\\
    & \sum_{i=1}^t x_i \geq \sum_{i=1}^t D_i\quad \forall i = 1, \dots, n - 1,\\
    & \sum_{i=1}^n x_i = \sum_{i=1}^n D_i,\\
    & 0 \leq x \leq M z,\\
    & z \in \{0,1\}^n.
  \end{align*}
  If $D_i = 0$, $i = 1, \dots, n-1$, $D_i = 1$, and $M = 1$, we obtain our problem.

>>>>

The methods in Section~\ref{sec:CoefficientTightening} cannot be used to
tighten bounds on each $y_k$ variable. The following lemma is valid, bu it
may cut off feasible solutions as shown in an example afterwards.

@@

@@

The following example shows that the lemma may indeed cut off feasible
solutions.

\begin{example}
  Consider the problem
  \begin{align*}
    \min\; &-y_1 \\
    \text{s.t.}\; &
    \begin{pmatrix} 2 & 0 \\ 0 & 2 \end{pmatrix} y_1
    + \begin{pmatrix} 3 & 0 \\ 0 & 3 \end{pmatrix} y_2
    + \begin{pmatrix} 1 & 0 \\ 0 & 1 \end{pmatrix} y_3
    - \begin{pmatrix} 1 & 0 \\ 0 & 1 \end{pmatrix} \succeq 0, \\
    & y_1 \in \{0,1\},\quad y_2 \in \{0,1\}, \quad y_3 \in [0,3]_\Z,
  \end{align*}
  which has optimal value $-1$. The scaling factor for the first variable
  is~$\mu_1 = \frac{1}{2}$, but changing the upper bound of~$y_1$
  to~$\mu_1$, i.e., fixing~$y_1 = 0$ cuts off all optimal
  solutions. However, the above lemma is not violated, since after
  fixing~$y_1 = \mu_1$, there still exists a feasible solution.
\end{example}

In order to correctly tighten variable bounds, assume that no coefficient
matrix~$A^k$ is indefinite, and that the variable
bounds~$\ell_k \leq y_k \leq u_k$ satisfy~$\ell_k >-\infty$
if~$A^k \preceq 0$ and~$u_k < \infty$ if~$A^k \succeq 0$ for
all~$k \in [m]$ . For an index~$k \in [m]$, define
\begin{align*}
  P_k \define \{i \neq k \suchthat A^i \succeq 0\},\qquad N_k \define \{i
  \neq k \suchthat A^i \preceq 0\},
\end{align*}
as well as
\begin{align*}
  \nu_k =
  \begin{cases}
    \min\, \{\nu \suchthat \nu A^k + \sum_{i\in P_k}A^i u_i + \sum_{j
      \in N_k} A^j \ell_j - A^0 \succeq 0 \}, & \text{if } A^k \succeq 0 \\
    \max\, \{\nu \suchthat \nu A^k + \sum_{i\in P_k}A^i u_i + \sum_{j
      \in N_k} A^j \ell_j - A^0 \succeq 0 \}, & \text{if } A^k \preceq 0.
  \end{cases}
\end{align*}

The following lemma shows that the lower or upper bounds of the variables
can be tightened, according to the semidefiniteness of the coefficient
matrices.
\begin{lemma}
  Let all~$A^k$, $k \in [m]$ be not indefinite, and let~$\ell_k >-\infty$
  if~$A^k \preceq 0$ as well as~$u_k < \infty$ if~$A^k \succeq 0$ for
  all~$k \in [m]$. Then, $\sum_{k=1}^m A^k y_k - A^0 \succeq 0$ implies
  \begin{align*}
    y_k
    \begin{cases}
      \geq \max \, \{\ell_k,\, \nu_k\}, & \text{if } A^k \succeq 0, \\
      \leq \min \, \{u_k,\, \nu_k\}, & \text{if } A^k \preceq 0.
    \end{cases}
  \end{align*}
\end{lemma}

\begin{proof}
  Let all~$A^k$, $k \in [m]$ be not indefinite, and let~$\ell_k >-\infty$
  if~$A^k \preceq 0$ as well as~$u_k < \infty$ if~$A^k \succeq 0$ for
  all~$k \in [m]$. Furthermore, let $\sum_{k=1}^m A^k y_k - A^0 \succeq 0$,
  and let~$k \in [m]$ with (w.l.o.g.)~$A^k \succeq 0$ and~$\nu_k \geq \ell_k$. Suppose
  that~$y_k < \nu_k$. Then, by definition of~$\nu_k$, there
  exists~$x \in \R^n$ with
  \begin{align*}
    0 &> x\T \Big(y_kA^k + \sum_{i \in P_k}A^i u_i + \sum_{i\in N_k}A^i
        \ell_i - A^0 \Big) x \\
      &= y_k x\T A^k x + \sum_{i \in P_k} u_i \underbrace{x\T A^i x}_{\geq 0} + \sum_{i\in N_k}
        \ell_i \underbrace{x\T A^i x}_{\leq 0} - x\T A^0 x \\
      &\geq y_k x\T A^k x + \sum_{i \in P_k} y_i x\T A^i x + \sum_{i\in N_k}
        y_i x\T A^i x - x\T A^0 x \\
      &= x\T \Big(\sum_{i=1}^m A^i y_i - A^0\Big) x,
  \end{align*}
  which is a contradiction to~$\sum_{k=1}^m A^k y_k - A^0 \succeq
  0$. Thus,~$y_k \geq \nu_k$.
\end{proof}

@@

\begin{proof}
  The constraint matrix of $S^k$ with respect to the equations and the
  inequalities $x - z \leq 0$ can be written as follows:
  \[
    M \define
    \begin{pmatrix}
      1 & \dots & 1 & 0 & \dots & 0 \\
      0 & \dots & 0 & 1 & \dots & 1 \\
      1 &       &   & -1 &      &   \\
        & \ddots&   &    & \ddots & \\
      0 &       & 1 & 0 &      & -1
    \end{pmatrix}.
  \]
  This matrix is totally unimodular. This can be seen by multiplying the
  first row with $-1$. Then we have exactly one $+1$ and $-1$ in each
  column. Thus, $M$ is the node-arc indicence matrix of a directed graph
  and hence TU. To take care of the variables bounds, one would add
  identity matrices to $M$, which do not change the property of being
  TU. Moreover, the right hand sides are integral and thus the polytope is
  integral.
\end{proof}

@@

  Then Aghezzaf and Wolsey consider lot sizing problems for given demands $D_1,
  \dots, D_n$:
  \begin{align*}
    \min\quad & c\T x + d\T z\\
    & \sum_{i=1}^t x_i \geq \sum_{i=1}^t D_i\quad \forall \, t = 1, \dots, n - 1,\\
    & \sum_{i=1}^n x_i = \sum_{i=1}^n D_i,\\
    & 0 \leq x \leq M z,\\
    & z \in \{0,1\}^n.
  \end{align*}
  If $D_i = 0$, $i = 1, \dots, n-1$, $D_n = 1$, and $M = 1$, we obtain our problem.
